Modeling and leveraging intuitive theories to improve vaccine attitudes

### 2022-06-06 paper/paper-papaja.Rmd
may-june revisions

authornote: |
  All data, and analysis code for this project are available at https://osf.io/sdfpj/. Preregistration information for Studies 3a and 3b is available at https://osf.io/csnez/registrations.

@@

Even before the coronavirus pandemic, misconceptions about vaccine safety drove public health crises around the world. Undervaccination resulted in outbreaks of measles affecting over one thousand children across the United States in 2019, and over 80,000 children across Europe in 2018 [@thornton2019]. Misconceptions about childhood vaccination---particularly the fraudulent claim that vaccines cause autism or other dangerous side-effects---have proved remarkably difficult to correct [e.g., @betsch.sachse2013; @horne.etal2015]. Sadly, only a handful of educational interventions have been found to improve vaccine attitudes [e.g., @dube.etal2015; @horne.etal2015; @vannice.etal2011; @wallace.etal2006]. This is especially  sobering in the context of the coronavirus pandemic: widespread reluctance to be vaccinated against COVID-19 threatens our ability to end a pandemic that has already claimed more than five million lives around the world.

The difficulty of convincing people to vaccinate themselves and their children illustrates a general pattern of findings showing that people often do not respond to evidence in ways that educators might wish [e.g. @lewandowsky.etal2012]. Part of the challenge in correcting misconceptions is that they are situated within larger, internally coherent belief systems---intuitive theories---that guide how people interpret and respond to evidence [e.g., @gershman2019; @lewandowsky.etal2012; @jern.etal2014]. 
Consider some of the beliefs people who are skeptical of childhood vaccination expressed when we asked them to explain their views in an open-ended qualitative survey (conducted pre-pandemic; _n_ = 16). Several participants saw vaccines as “unnatural” and expressed concerns about toxic additives in vaccines, such as heavy metals and mercury. One said, “Vaccines are unnatural additions to the body. The human body should naturally fight off the disease and we should not be controlling how nature plays its role on mankind.” Some doubted whether certain diseases were worth worrying over, while others feared that the contents of vaccines “are too strong and too concentrated for a baby’s body to properly handle.” A number of respondents were skeptical of the medical community’s objectivity in testing the safety of vaccines. One respondent wrote that vaccines were “A BIG MONEY MAKING SCAM BY THE BIG PHARMS” (emphasis theirs); another thought that health professionals “are brainwashed by the medical media and the government.”

@@

We argue that correcting misconceptions requires presenting evidence in a way that is sensitive to and persuasive within the broader conceptual context in which those misconceptions reside [see @gershman2019; @weisman.markman2017]. Understanding the revision of any one specific belief (e.g., a misconception like *vaccines are dangerous*) demands a holistic understanding of the wider set of interrelated beliefs that make up people's intuitive theories.

@@

Fortunately, the formalism of Bayesian networks provides tools for navigating these challenges. Roughly, when the relationships among beliefs are unknown, we can use the qualitative features of intuitive theories to *infer* the structure of an intuitive theory from observed correlations among beliefs. Structure learning algorithms [see @scutari.denis2021] provide a means for searching through the (super-exponentially large) possible space of Bayesian networks to find the structure most likely to have generated some observed data. These approaches allow us to scale the Bayesian network modeling framework beyond the smaller models characteristic of early computational work on intuitive theories [e.g., @griffiths.tenenbaum2005].

@@

We translated these qualitative insights into 14 psychometrically-valid scales. In addition to 1) _vaccination intentions_ (e.g. “If I had a young baby I would have him or her vaccinated.”), these included a variety of beliefs about vaccines, including beliefs about 2) _vaccine danger_, the overall danger of vaccines (e.g., “Many children have severe adverse reactions to the MMR vaccine”); 3) _toxic additives in vaccines_ (e.g., “Exposure to certain additives in vaccines can cause major health problems.”); and 4) _vaccine effectiveness_, how effective vaccines are in preventing disease (e.g., “Your chances of getting a disease after being vaccinated against it are incredibly low”); as well as a variety of specific claims about childhood diseases like measles, mumps, and rubella, including beliefs about 5) _disease rarity _(e.g., “Diseases like measles and whooping cough are so rare that there is no real need to worry about them”) and 6) _disease severity_ (e.g., “Diseases like measles and whooping cough are extremely dangerous for young children”). Beyond this, we theorized that participants’ understanding of the infant immune system might be relevant, including beliefs that 7) the infant immune system is weak (_IIS: weakness_; e.g., “Babies’ immune systems are weak”); 8) the infant immune system is limited in its capacity and can be easily overwhelmed (_IIS: limited capacity_; e.g., “Babies’ immune systems are severely limited in how much they can cope with at one time.”); and 9) vaccines strain the infant immune system (_IIS: vaccines strain_; e.g., “Vaccines strain the immune system much like the actual disease”). We also included general theories about parenting and medicine: 10) general _parental protectiveness_ (e.g., “Parents should strive to control their baby’s environments as much as possible”); 11) _parental expertise_, namely the belief that parents usually know more about their children’s health than medical experts (e.g., “Every individual is different and a parent knows their own child best”); and 12) _medical skepticism_, including concerns about pharmaceutical companies and corruption in the medical community (e.g., “Pharmaceutical companies put pressure on the FDA and CDC to suppress negative findings”); and two two broad worldviews: 13) _naturalism_, a general preference for natural over artificial things (e.g., “Natural things are always better than synthetic alternatives”); and 14) _holistic balance_, one important aspect of attitudes toward alternative medicine (e.g., “The body is essentially self-healing and the task of a health care provider is to assist the healing process”; [@mcfadden.etal2010]).

Each scale measured agreement with 4-6 statements, including at least one reverse-coded item, and was highly reliable (observed reliability, Study 1: all Cronbach’s $\alpha \geq .73$ .73). We developed 13 of these scales, and included one preexisting scale (the “holistic balance” subscale from [@mcfadden.etal2010]. See SI for all items included in all scales.

@@

$$P(y) = \Big(1-(1-w_0)\prod_i(1-w_i)^{x_i}\Big)\prod_j(1-w_j)^{x_j}$$

@@

Having transformed participants’ scores on each belief scale to fall between 0 and 1, we estimated the parameters of the model as a beta regression using maximum likelihood methods (see SI).

@@

In particular, we applied a constraint we called the “generative model principle,” based on the assumption that the intuitive theory is a generative model, so that these influence relations (“edges”) should flow from causes or other “generating” states of affairs toward effects. For instance, we stipulated that vaccine intentions---our outcome of interest---would be a “child node,” with edges only flowing inwards and not outwards (since one’s intention to vaccinate cannot “generate” the facts about vaccinations safety, effectiveness, etc). This generative model principle instills a common core assumption in research on intuitive theories and causal explanatory reasoning[e.g. @gerstenberg.etal2021; @gopnik.etal2004; @cheng1997].

@@

To avoid overfitting, we used an 80% split of the data (*n* = 904) for training, and reserved the remaining 20% for evaluating model performance. Applying these techniques resulted in an estimate of the network shown in Figure 1.

```{r fig1-maingraph, fig.align='center', out.width="90%", fig.cap="Bayesian network model resulting from hill-climbing algorithm with our generative-model constraints applied to data in Study 1 training split (n = 904). The network has 29 edges connecting these 14 variables. Edge polarity and weights are indicated by color, with red hues indicating negative or preventive influences, and blue hues indicating positive or generating influences. The direction of the edge was not determinable from the data and modeling constraints for two edges: $\\text{Naturalism} \\rightarrow \\text{Holistic Balance}$ and $\\text{Medical Skepticism} \\rightarrow \\text{Parental Expertise}$; see SI."}

@@

We evaluated the model’s performance by testing its accuracy in predicting responses in the 20% testing split (*n* = 226). We generated predictions for held-out participants’ responses for each variable by conditioning the network on the remaining 13 observed variables. Figure 2 compares the model’s predictions with participants’ actual responses. Collapsing across all variables, the average correlation between predicted and observed responses was *r* = `r round(avg_corr, 3)`, accounting for `r round(100*avg_corr^2, 0)`% of the variance in observed responses.

```{r fig2-validation, fig.align='center', out.width="100%", fig.cap="Observed responses and model predictions for each belief among participants in the Study 1 testing split. Correlations between observed and predicted values ranged from .380 to .898 across the different belief scales. In general, the model showed greater predictive accuracy for more central beliefs (e.g., \\emph{vaccine danger}) than for more peripheral beliefs (e.g., \\emph{parental protectiveness}; see SI for details)."}
knitr::include_graphics("../local/valid_plt.png")
```

What does this model reveal about the intuitive theory underlying vaccination decisions?

@@

In other cases, our data and explicit constraints do not perfectly identify the directions of the edges. This is the case for two edges (Naturalism $\rightarrow$  Holistic Balance and Medical Skepticism $\rightarrow$ Parental Expertise; see SI for details). In both cases, we do not feel it is intuitive which direction the edge should flow in, nor even that it should be directed at all; we might even speculate there is an additional unmeasured variable that connects these beliefs and that would best account for their association.

@@

We conducted a mixed-effects Bayesian ordinal regression to assess the effects of the disease risk intervention on responses to the *Vaccination Intentions* scale items, using the “brms” package for R [@burkner2017]. We regressed responses on phase (Phase 1 vs. 2, dummy-coded with Phase 1 as the baseline), condition (dummy-coded with the disease risk condition as the baseline, so as to assess difference scores in our condition of primary interest), and the interaction between phase and condition. We included random intercepts by subject and scale item. In this model, the unique effect of the intervention on vaccination intention scores is assessed by the interaction term (indicating the unique effect of Phase 2 for intervention versus no-intervention participants). The disease risk intervention increased participants’ vaccination intention ratings (interaction effect: $b$ = `r round(s2_reg$fixed[9,1],3)`, 95% CI: [`r round(s2_reg$fixed[9,3], 3)`, `r round(s2_reg$fixed[9,4], 3)`]). We consider this to be a full replication of Horne and colleagues' [-@horne.etal2015] results. See Figure 3, panel A.

@@

```{r fig3-s2, fig.align='center', out.width="100%", fig.cap="Belief changes following the intervention in Study 2. Left, mean difference scores from pretest to posttest for the 14 belief scales across experimental conditions. Right, mean observed and model predicted change scores for the disease risk intervention condition. In both plots, error bars represent 95\\% confidence intervals. In both plots, changes were calculated after transforming responses to lie between 0 and 1."}

@@

Where *x* was a binary variable representing the presence (1) or absence (0) of evidence at posttest, and participants’ prior and posterior (pretest and posttest) credences were transformed into log odds. This allowed us to estimate the log evidence ratio implied by participants’ reactions to the intervention. Then, we constructed a CPT consistent with that evidence ratio.

Using this augmented model, we are able to “condition” the model on this evidence node, and thereby predict how observing evidence will affect all of the beliefs in the network.  As shown in Figure 3, panel B, this simulation is a reasonable approximation of what we observed in Study 2: The model’s predictions correlate strongly with observed changes (*r* = `r results$s2$modcorr`).

Given that our cognitive model is parameterized via noisy-logical functions, it should be noted that the marginal predicted influence of each belief is closely related to its first-order correlation with disease severity. This is in keeping with the qualitative features of intuitive theories we laid out in the introduction, and we should expect this to hold true (on average) for any plausible model of an intuitive theory. However, future models with more complex or nuanced parameterizations could potentially make additional predictions not evident from first-order correlations, such as individualized predictions for how individuals or groups would respond to evidence.

@@

One advantage of having developed a formal, computational representation of a network of beliefs supporting vaccination decisions is that it can guide our search for effective interventions, allowing us to compare how interventions targeting each of these other beliefs might potentially change people’s vaccination intentions (see Figure 4). Countering myths about toxic additives is one approach the model developed in Study 1 predicts should be effective.

```{r fig4-predimpacts, fig.align='center',  out.width="50%", fig.cap="Predicted change in vaccination intentions following an equivalent hypothetical change in each of the other 13 beliefs."}
knitr::include_graphics("../local/pred_change_plt.png")
```

@@

We conducted two preregistered experiments[^prereg] (Study 3a and Study 3b) testing the effects of this novel intervention aimed at dispelling concerns about toxic additives in vaccines using the same pretest-posttest design as Study 2. These studies followed similar design procedures as Study 2, with Phase 2 conducted 1-2 days after Phase 1. In both Study 3a (April 2019) and Study 3b (May 2019; combined final *n* = `r results$s3$demo$n_complete`) the vaccine ingredients intervention was compared against no-intervention control. Unlike Study 2, we did not stratify participants before random assignment to conditions.

@@

We recruited `r results$s3$demo$n_recruited` people to participate in a two-day study via mTurk in April (Experiment 3A) and June (Experiment 3B) of 2019. One day after their initial participation, participants who completed part 1 and passed attention checks were invited via email to participate in Day 2 of the study, with `r results$s3$demo$n_recruited_post` participants returning (`r results$s3$demo$n_eligible` invited, retaining `r round(results$s3$demo$n_recruited_post/results$s3$demo$n_eligible*100,1)`% of the Day 1 sample). Participants were paid \$1.35 for their participation in each phase of the study and were allowed to participate in part 2 up to two days after their initial participation in part 1. `r results$s3$demo$n_failed_post` participants (`r round( (results$s3$demo$n_failed_post/results$s3$demo$n_recruited_post)*100,1)`%) failed at least one attention check on Day 2 and were excluded from further analysis.  This left a final sample of `r results$s3$demo$n_complete` participants. See SI for demographic information about this sample.

@@

As they shared the same design, we combined data from Studies 3a and 3b for analysis (but see SI for separate analyses by study). As in Study 2, we conducted a mixed-effects Bayesian ordinal regression to assess the effects of the vaccine ingredients intervention on participants’ *vaccination intentions* scores. The vaccine ingredients intervention improved participant’s vaccination intention scores relative to the control conditions (interaction effect: $b$ = `r round(s3_reg$fixed[9,1],3)`, 95% CI: [`r round(s3_reg$fixed[9,3], 3)`, `r round(s3_reg$fixed[9,4], 3)`]); see Figure 5, panel A. It is worth noting that reliable changes were also observed from pretest to posttest in the control conditions (phase effect: $b$ = `r round(s3_reg$fixed[7,1],3)`, 95% CI: [`r round(s3_reg$fixed[7,3], 3)`, `r round(s3_reg$fixed[7,4], 3)`]), suggesting that these attitudes were also being influenced by other factors during the course of the study. We discuss this further below, but given that these studies were randomized controlled trial designs, these findings do not impair our ability to conclude that the intervention was causally efficacious in affecting vaccine intentions.

@@

To model the effect of the vaccine ingredients intervention, we followed the same procedure as in Study 2, but this time augmented the model with a node representing the vaccine ingredients intervention as a child of the vaccine toxins belief node (for details, see SI). This allowed us to predict how other beliefs in the network would change in light of the observed changes participant’s beliefs about whether vaccines contain dangerous toxins; see Figure 5, panel B.

@@

```{r fig5-s3, fig.align='center',  out.width="100%", fig.cap="Belief changes following the intervention in studies 3a and 3b. Left, mean difference scores from pretest to posttest for the 14 belief scales across experimental conditions. Right, mean observed and model predicted change scores for the disease risk intervention condition. In both plots, changes were calculated after combining data from both studies and transforming responses to lie between 0 and 1 and error bars represent 95\\% confidence intervals."}
knitr::include_graphics("../local/s3_plt.png")
```

@@

```{r fig6-s4, fig.align='center',  out.width="100%", fig.cap="Participants’ belief changes from participation in Study 1 to Study 4. Left, mean difference scores comparing beliefs reported in August-September 2017 and May 2019 for the 14 belief scales, across participants who did and did not report exposure to news of the ongoing New York measles outbreak. Right, mean observed and model predicted change scores for those participants who did report exposure to this news. Belief changes were modeled assuming that news of the outbreak directly impacted participant's vaccine intentions, the most central of the 14 beliefs, in order to evaluate the overall coherence of these belief changes. In both plots, changes were calculated after transforming responses to lie between 0 and 1 and error bars represent 95\\% confidence intervals."}
knitr::include_graphics("../local/s4_plt.png")
```

@@

Were changes in participants' beliefs following exposure to news of the measles outbreak coherent with respect to the intuitive theory specified by the model developed in Study 1? Unlike in the intervention studies, where a specific belief was targeted by the intervention, it is likely that many beliefs were addressed by media coverage of these outbreaks. To assess whether these changes were coherent, we examined how they were related to the most central of the beliefs we measured, *vaccine intentions*, using our model to predict how the 13 other beliefs in the intuitive theory should be expected to change given the average magnitude of the change observed in *vaccine intentions.* That is, we followed the same procedure as in previous studies, but here augmented the model with a node representing media exposure as a child of the *vaccine intentions* node.

@@

@@

>>>>

authornote: |
  All data, and analysis code for this project are available at https://osf.io/sdfpj/. Preregistration information for Studies 3a and 3b is available at https://osf.io/csnez/registrations. Earlier modeling and analyses of data from Study 1 were presented at the 40th Annual Conference of the Cognitive Science Society. A six-page paper also appeared in the proceedings [@powell.etal2018].

@@

Even before the coronavirus pandemic, misconceptions about vaccine safety drove public health crises around the world. Undervaccination resulted in outbreaks of measles affecting over one thousand children across the United States in 2019, and over 80,000 children across Europe in 2018 [@thornton2019]. Misconceptions about childhood vaccination---particularly the fraudulent claim that vaccines cause autism or other dangerous side-effects---have proved remarkably difficult to correct [e.g., @betsch.sachse2013; @horne.etal2015; @lewandowsky.etal2012]. Sadly, only a handful of educational interventions have been found to improve vaccine attitudes [e.g., @dube.etal2015; @horne.etal2015; @vannice.etal2011; @wallace.etal2006]. This is especially  sobering in the context of the coronavirus pandemic: widespread reluctance to be vaccinated against COVID-19 threatens our ability to end a pandemic that has already claimed more than five million lives around the world.

The difficulty of convincing people to vaccinate themselves and their children illustrates a general pattern of findings showing that people often do not respond to evidence in ways that educators might wish [e.g. @lewandowsky.etal2012]. Part of the challenge in correcting misconceptions is that they are situated within larger, internally coherent belief systems---intuitive theories---that guide how people interpret and respond to evidence [e.g., @gershman2019; @lewandowsky.etal2012; @jern.etal2014; @cook.lewandowsky2016].

Consider some of the beliefs people who are skeptical of childhood vaccination expressed when we asked them to explain their views in an open-ended qualitative survey (conducted pre-pandemic; _n_ = 16). Several participants saw vaccines as “unnatural” and expressed concerns about toxic additives in vaccines, such as heavy metals and mercury. One said, “Vaccines are unnatural additions to the body. The human body should naturally fight off the disease and we should not be controlling how nature plays its role on mankind.” Some doubted whether certain diseases were worth worrying over, while others feared that the contents of vaccines “are too strong and too concentrated for a baby’s body to properly handle.” A number of respondents were skeptical of the medical community’s objectivity in testing the safety of vaccines. One respondent wrote that vaccines were “A BIG MONEY MAKING SCAM BY THE BIG PHARMS” (emphasis theirs); another thought that health professionals “are brainwashed by the medical media and the government.”

@@

We argue that correcting misconceptions requires presenting evidence in a way that is sensitive to and persuasive within the broader conceptual context in which those misconceptions reside [see @gershman2019; @weisman.markman2017; @cook.lewandowsky2016]. Understanding the revision of any one specific belief (e.g., a misconception like *vaccines are dangerous*) demands a holistic understanding of the wider set of interrelated beliefs that make up people's intuitive theories.

@@

Fortunately, the formalism of Bayesian networks provides tools for navigating these challenges. Roughly, when the relationships among beliefs are unknown, we can use the qualitative features of intuitive theories to *infer* the structure of an intuitive theory from observed correlations among beliefs. Structure learning algorithms [see @scutari.denis2021] provide a means for searching through the (super-exponentially large) possible space of Bayesian networks to find the structure most likely to have generated some observed data. These approaches allow us to scale the Bayesian network modeling framework beyond the smaller models characteristic of early computational work on intuitive theories [e.g., @cook.lewandowsky2016; @griffiths.tenenbaum2005; @jern.etal2012].

@@

We translated these qualitative insights into 14 psychometrically-valid scales. In addition to 1) _vaccination intentions_ (e.g. “If I had a young baby I would have him or her vaccinated.”), these included a variety of beliefs about vaccines, including beliefs about 2) _vaccine danger_, the overall danger of vaccines (e.g., “Many children have severe adverse reactions to the MMR vaccine”); 3) _toxic additives in vaccines_ (e.g., “Exposure to certain additives in vaccines can cause major health problems.”); and 4) _vaccine effectiveness_, how effective vaccines are in preventing disease (e.g., “Your chances of getting a disease after being vaccinated against it are incredibly low”); as well as a variety of specific claims about childhood diseases like measles, mumps, and rubella, including beliefs about 5) _disease rarity_ (e.g., “Diseases like measles and whooping cough are so rare that there is no real need to worry about them”) and 6) _disease severity_ (e.g., “Diseases like measles and whooping cough are extremely dangerous for young children”). Beyond this, we theorized that participants’ understanding of the infant immune system might be relevant, including beliefs that 7) the infant immune system is weak (_IIS: weakness_; e.g., “Babies’ immune systems are weak”); 8) the infant immune system is limited in its capacity and can be easily overwhelmed (_IIS: limited capacity_; e.g., “Babies’ immune systems are severely limited in how much they can cope with at one time.”); and 9) vaccines strain the infant immune system (_IIS: vaccines strain_; e.g., “Vaccines strain the immune system much like the actual disease”). We also included general theories about parenting and medicine: 10) general _parental protectiveness_ (e.g., “Parents should strive to control their baby’s environments as much as possible”); 11) _parental expertise_, namely the belief that parents usually know more about their children’s health than medical experts (e.g., “Every individual is different and a parent knows their own child best”); and 12) _medical skepticism_, including concerns about pharmaceutical companies and corruption in the medical community (e.g., “Pharmaceutical companies put pressure on the FDA and CDC to suppress negative findings”); and two two broad worldviews: 13) _naturalism_, a general preference for natural over artificial things (e.g., “Natural things are always better than synthetic alternatives”); and 14) _holistic balance_, one important aspect of attitudes toward alternative medicine (e.g., “The body is essentially self-healing and the task of a health care provider is to assist the healing process”; [@mcfadden.etal2010]).

Each scale measured agreement with 4-6 statements, including at least one reverse-coded item, and was highly reliable (observed reliability, Study 1: all Cronbach’s $\alpha \geq .73$). We developed 13 of these scales, and included one preexisting scale (the “holistic balance” subscale from [@mcfadden.etal2010]. See SI for all items included in all scales.

@@

$$P(y=1) = \Big(1-(1-w_0)\prod_i(1-w_i)^{x_i}\Big)\prod_j(1-w_j)^{x_j}$$

@@

A cognitive model composed of these types of relations is capable of capturing a causal system of non-interactive generative and preventive causes. But it bears noting that this simplification comes with limitations. For instance, returning to the example network in Figure 1d, this parameterization would not permit capturing the relationships between dryness, lightning, and fire. Dryness is an enabling condition that both increases the probability of fire but also increases the probability that lightning would cause a fire. So the influences of these variables are not independent and would need a more complex parameterization to be properly represented.

Finally, having transformed participants’ scores on each belief scale to fall between 0 and 1, we estimated the parameters of this model as a Beta regression using maximum likelihood methods (see SI for details).

@@

In particular, we applied a constraint we called the “generative model principle,” based on the assumption that the intuitive theory is a generative model, so that these influence relations (“edges”) should flow from causes or other “generating” states of affairs toward effects. For instance, we stipulated that vaccine intentions---our outcome of interest---would be a “child node,” with edges only flowing inwards and not outwards (since one’s intention to vaccinate cannot “generate” the facts about vaccinations safety, effectiveness, etc). This generative model principle instills a common core assumption in research on intuitive theories and causal explanatory reasoning [e.g. @gerstenberg.etal2021; @gopnik.etal2004; @cheng1997].

@@

To avoid overfitting, we used an 80% split of the data (*n* = 904) for training, and reserved the remaining 20% for evaluating model performance. We compared performance of different structure learning algorithms using k-fold cross validation within the training split. Based on these results, we chose to apply a hill-climbing learning algorithm using our generative model constraints (see SI for details). Applying these techniques resulted in an estimate of the network shown in Figure 1.

```{r fig1-maingraph, fig.align='center', out.width="90%", fig.cap="Bayesian network cognitive model resulting from application of hill-climbing algorithm with our generative-model constraints applied to data in Study 1 training split (n = 904). The network has 29 edges connecting these 14 variables. Edge polarity and weights are indicated by color, with red hues indicating negative or preventive influences, and blue hues indicating positive or generating influences. The direction of the edge was not determinable from the data and modeling constraints for two edges: $\\text{Naturalism} \\rightarrow \\text{Holistic Balance}$ and $\\text{Medical Skepticism} \\rightarrow \\text{Parental Expertise}$; see SI."}

@@

We evaluated the resulting cognitive model’s performance by testing its accuracy in predicting responses in the 20% testing split (*n* = 226). We generated predictions for held-out participants’ responses for each variable by conditioning the network on the remaining 13 observed variables. Figure 2 compares the model’s predictions with participants’ actual responses. Collapsing across all variables, the average correlation between predicted and observed responses was *r* = `r round(avg_corr, 3)`, accounting for `r round(100*avg_corr^2, 0)`% of the variance in observed responses.

Our goal in developing this cognitive model was to create a psychologically plausible and intelligible model of the intuitive theory surrounding vaccination decisions. However, it is also reasonable to ask whether other less-constrained models might be more predictive. To address this, we compared the Bayesian network model’s predictions to a suite of 14 naive “maximal” regression models, each using values on all 13 other belief scales as predictors for each belief. Our simpler model provided equally strong predictions (see SI). And, as will be evidenced in later studies, our model can also be used to generate accurate predictions about how beliefs will change following evidence.

```{r fig2-validation, fig.align='center', out.width="100%", fig.cap="Observed responses and cognitive model predictions for each belief among participants in the Study 1 testing split. Correlations between observed and predicted values ranged from .380 to .898 across the different belief scales. In general, the model showed greater predictive accuracy for more central beliefs (e.g., \\emph{vaccine danger}) than for more peripheral beliefs (e.g., \\emph{parental protectiveness}; see SI for details)."}
knitr::include_graphics("../local/valid_plt.png")
```

What does this cognitive model reveal about the intuitive theory underlying vaccination decisions?

@@

In other cases, our data and explicit constraints do not perfectly identify the directions of the edges. This is the case for two edges (Naturalism $\rightarrow$  Holistic Balance and Medical Skepticism $\rightarrow$ Parental Expertise; see SI for details). In both cases, we do not feel it is intuitive which direction the edge should flow in, nor even that it should be directed at all. Such a counterintuitive result might be taken as indicating a limitation of Bayesian network models, which are composed solely of directed, acyclic relationships. Prior work has successfully modeled coherent belief revision by leveraging models with undirected or bidirected relations [e.g. @spellman.etal1993]. Alternatively, we might imagine there is an additional unmeasured variable that connects these beliefs that could more intelligibly account for their association within the context of a directed network. For instance, a more abstract epistemic belief might influence both medical skepticism and parental expertise beliefs. To speculate, such an abstract belief might be in one's own expertise or in how we come to have knowledge (e.g. through direct experience versus science).

@@

We conducted a mixed-effects Bayesian ordinal regression to assess the effects of the disease risk intervention on responses to the *Vaccination Intentions* scale items, using the “brms” package for R [@burkner2017]. We regressed responses on phase (Phase 1 vs. 2, dummy-coded with Phase 1 as the baseline), condition (dummy-coded with the disease risk condition as the baseline, so as to assess difference scores in our condition of primary interest), and the interaction between phase and condition. We included random intercepts by subject and scale item. In this model, the unique effect of the intervention on vaccination intention scores is assessed by the interaction term (indicating the unique effect of Phase 2 for intervention versus no-intervention participants). The disease risk intervention increased participants’ vaccination intention ratings (interaction effect: $b$ = `r round(s2_reg$fixed[9,1],3)`, 95% Credible Interval (CI): [`r round(s2_reg$fixed[9,3], 3)`, `r round(s2_reg$fixed[9,4], 3)`]). We consider this to be a full replication of Horne and colleagues' [-@horne.etal2015] results. See Figure 3, panel A.

@@

```{r fig3-s2, fig.align='center', out.width="100%", fig.cap="Belief changes following the intervention in Study 2. Left, mean difference scores from pretest to posttest for the 14 belief scales across experimental conditions. Right, mean observed and predicted change scores from the cognitive model for the disease risk intervention condition. In both plots, error bars represent 95\\% confidence intervals. In both plots, changes were calculated after transforming responses to lie between 0 and 1."}

@@

Where *x* was a binary variable representing the presence (1) or absence (0) of evidence at posttest, and participants’ prior and posterior (pretest and posttest) credences were transformed into log odds. The variables $ER_0$ and $ER_1$ represent measures of the evidence seen by participants in the control and intervention conditions. These are defined as:

$$ER_i = \ln\frac{(P(d_i|h=1))}{(P(d_i|h=0))}$$

In equation X, $ER_0$ represents whatever background "evidence" participants would see between phases outside  the experimental context, that would be shared by participants in both conditions. $ER_1$ represents the evidence provided by the intervention information, which is assumed to be independent from $ER_0$. These log evidence ratios were both estimated via regression. Then, we constructed a conditional probability table consistent with $ER_1$ and augmented the cognitive model with this evidence node (see SI for details).

Using this augmented model, we are able to “condition” our cognitive model on this evidence node, and thereby predict how observing evidence will affect all of the beliefs in the network.  As shown in Figure 3, panel B, this simulation is a reasonable approximation of what we observed in Study 2: The model’s predictions correlate strongly with observed changes (*r* = `r results$s2$modcorr`).

Given that our cognitive model is parameterized via noisy-logical functions wherein each parent independently influences each child node, it should be noted that the marginal predicted influence of each belief is closely related to its first-order correlation with disease severity. This is in keeping with the qualitative features of intuitive theories we laid out in the introduction, and we should often expect this to hold true on average for any plausible model of an intuitive theory. However, future models with more complex or nuanced parameterizations could potentially make additional predictions not evident from first-order correlations, such as individualized predictions for how individuals or groups would respond to evidence.

@@

One advantage of having developed a formal, computational representation of a network of beliefs supporting vaccination decisions is that it can guide our search for effective interventions, allowing us to compare how interventions targeting each of these other beliefs might potentially change people’s vaccination intentions (see Figure 4). Countering myths about toxic additives is one approach the cognitive model developed in Study 1 predicts should be effective.

```{r fig4-predimpacts, fig.align='center',  out.width="50%", fig.cap="Predicted change in vaccination intentions based on cognitive model following an equivalent hypothetical change in each of the other 13 beliefs."}
knitr::include_graphics("../local/pred_change_plt.png")
```

@@

We conducted two preregistered experiments[^prereg] (Study 3a and Study 3b) testing the effects of this novel intervention aimed at dispelling concerns about toxic additives in vaccines using the same pretest-posttest design as Study 2. These studies followed similar design procedures as Study 2, with Phase 2 conducted 1-2 days after Phase 1. In both Study 3a (April 2019) and Study 3b (June 2019; combined final *n* = `r results$s3$demo$n_complete`) the vaccine ingredients intervention was compared against no-intervention control. Unlike Study 2, we did not stratify participants before random assignment to conditions.

@@

We recruited `r results$s3$demo$n_recruited` people to participate in a two-day study via mTurk in April (Experiment 3a) and June (Experiment 3b) of 2019. One day after their initial participation, participants who completed part 1 and passed attention checks were invited via email to participate in Day 2 of the study, with `r results$s3$demo$n_recruited_post` participants returning (`r results$s3$demo$n_eligible` invited, retaining `r round(results$s3$demo$n_recruited_post/results$s3$demo$n_eligible*100,1)`% of the Day 1 sample). Participants were paid \$1.35 for their participation in each phase of the study and were allowed to participate in part 2 up to two days after their initial participation in part 1. `r results$s3$demo$n_failed_post` participants (`r round( (results$s3$demo$n_failed_post/results$s3$demo$n_recruited_post)*100,1)`%) failed at least one attention check on Day 2 and were excluded from further analysis.  This left a final sample of `r results$s3$demo$n_complete` participants. See SI for demographic information about this sample.

@@

As they shared the same design, we combined data from Studies 3a and 3b for analysis (but see SI for separate analyses by study). As in Study 2, we conducted a mixed-effects Bayesian ordinal regression to assess the effects of the vaccine ingredients intervention on participants’ *vaccination intentions* scores. The vaccine ingredients intervention improved participants' vaccination intention scores relative to the control conditions (interaction effect: $b$ = `r round(s3_reg$fixed[9,1],3)`, 95% CI: [`r round(s3_reg$fixed[9,3], 3)`, `r round(s3_reg$fixed[9,4], 3)`]); see Figure 5, panel A. It is worth noting that reliable changes were also observed from pretest to posttest in the control conditions (phase effect: $b$ = `r round(s3_reg$fixed[7,1],3)`, 95% CI: [`r round(s3_reg$fixed[7,3], 3)`, `r round(s3_reg$fixed[7,4], 3)`]), suggesting that these attitudes were also being influenced by other factors during the course of the study. We discuss this further below, but given that these studies were randomized controlled trial designs, these findings do not impair our ability to conclude that the intervention was causally efficacious in affecting vaccine intentions.

@@

To model the effect of the vaccine ingredients intervention, we followed the same procedure as in Study 2, but this time augmented the cognitive model with a node representing the vaccine ingredients intervention as a child of the vaccine toxins belief node (for details, see SI). This allowed us to predict how other beliefs in the network would change in light of the observed changes in participant’s beliefs about whether vaccines contain dangerous toxins; see Figure 5, panel B.

@@

```{r fig5-s3, fig.align='center',  out.width="100%", fig.cap="Belief changes following the intervention in studies 3a and 3b. Left, mean difference scores from pretest to posttest for the 14 belief scales across experimental conditions. Right, mean observed and  predicted change scores from the cognitive model for the disease risk intervention condition. In both plots, changes were calculated after combining data from both studies and transforming responses to lie between 0 and 1 and error bars represent 95\\% confidence intervals."}
knitr::include_graphics("../local/s3_plt.png")
```

@@

```{r fig6-s4, fig.align='center',  out.width="100%", fig.cap="Participants’ belief changes from participation in Study 1 to Study 4. Left, mean difference scores comparing beliefs reported in August-September 2017 and May 2019 for the 14 belief scales, across participants who did and did not report exposure to news of the ongoing New York measles outbreak. Right, mean observed and  predicted change scores from the cognitive model for those participants who did report exposure to this news. Belief changes were modeled assuming that news of the outbreak directly impacted participant's vaccine intentions, the most central of the 14 beliefs, in order to evaluate the overall coherence of these belief changes. In both plots, changes were calculated after transforming responses to lie between 0 and 1 and error bars represent 95\\% confidence intervals."}
knitr::include_graphics("../local/s4_plt.png")
```

@@

Were changes in participants' beliefs following exposure to news of the measles outbreak coherent with respect to the intuitive theory specified by the cognitive model developed in Study 1? Unlike in the intervention studies, where a specific belief was quite precisely targeted by the intervention, it is likely that many beliefs were addressed by media coverage of these outbreaks. Given that we cannot reasonably measure the exact information that participants were exposed to, it is not feasible to attempt to precisely model the impacts of this total evidence. However, we can leverage our model to assess whether the changes in their beliefs were *coherent*.

Throughout our prior studies, we have focused primarily on *vaccine intentions* as the most crucial and proximal belief for determining vaccination decisions. So, we sought to examine whether the changes across all the measured beliefs were coherent with respect to the changes in vaccination intensions. Using our model, we generated predictions for how the 13 other beliefs in the intuitive theory should be expected to change given the average magnitude of the change observed in vaccine intentions. To accomplish this, we followed the same procedure as in previous studies, but here augmented the model with a node representing media exposure as a child of the *vaccine intentions* node.

@@

[talk about magnitude of the effects. tailoring could strengthen. timing also matters, in context of physician visit]

@@

## Broader implications and applications

[discuss broader implications / applications as requested]
